\section{Introduction}
Robots that climb or follow staircases and assistive devices for the
visually impaired must reason about three-dimensional structure to act
safely. Legged and wheeled platforms increasingly rely on egocentric
depth sensing to negotiate stairs and other discontinuous terrain
\cite{agarwal2022legged,belter2016adaptive,stelzer2012stereo}, while
shared-autonomy systems demand reliable geometric cues before a foot or
wheel is committed to a surface \cite{marion2017director}. In all of
these settings the central perceptual question is the same: which parts
of an observed point cloud constitute a \emph{tread}, the steppable
horizontal surface on which weight can be safely placed, as opposed to a
near-vertical \emph{riser} or spurious edge and outlier returns.

This problem is safety-critical. A tread mislabeled as non-steppable
merely sacrifices an option, but a riser or edge mislabeled as steppable
invites a foothold onto a vertical face, producing slips, stumbles, or
falls. The same hazard motivates stair perception for the visually
impaired, where undetected steps are a recognised fall hazard that motivates
dedicated stair-detection systems \cite{matsumura2022deep}. Reliable tread detection therefore requires not
only high recall but also high precision on the steppable class.

Recent work approaches stair and terrain perception with deep neural
networks that learn point-wise semantics directly from raw clouds
\cite{qi2016pointnet,matsumura2022deep}. Such methods are powerful but
data hungry: they require large volumes of densely labeled
three-dimensional scans, which are costly to acquire for the specific
geometry of stairs, and they offer limited interpretability. Equally
problematic, the stair tread sub-problem lacks controlled, reproducible
benchmarks with exact ground-truth labels, so reported results are
difficult to compare across studies. Classical geometric methods, by
contrast, are fast, transparent, and label-free, yet despite their
maturity in plane segmentation and traversability estimation
\cite{chavezgarcia2017learning,gomes2023survey} they remain
under-compared on the staircase setting under matched conditions.

We address this gap with a controlled, fully reproducible study of
classical geometric segmentation for steppable-surface detection on
staircases. Our contributions are as follows:
\begin{itemize}
  \item A parametric, fully labeled staircase point-cloud generator that
        emits exact tread, riser, and other labels while modeling LiDAR
        range jitter and outliers, enabling reproducible evaluation.
  \item A unified three-class problem formulation together with the
        implementation of four classical methods under a common
        orientation-aware pipeline: RANSAC plane fitting, PCA normal
        analysis, DBSCAN with slope confirmation, and a height-histogram
        level detector.
  \item A quantitative comparison of all four methods on the steppable
        (tread) class, accompanied by a controlled noise-sensitivity
        study that characterizes degradation as height noise grows.
  \item A safety-oriented discussion of the results for robotic stair
        traversal, relating precision and intersection-over-union to the
        risk of unsafe footholds.
\end{itemize}

The remainder of this paper is organized as follows: Section~II reviews
related work, Section~III presents the proposed data generator and the
four segmentation methods, Section~IV reports the comparative evaluation
and noise study, and Section~V concludes.
