\section{Proposed Method}
\label{sec:method}

This section formalises the steppable-surface segmentation problem, describes a
parametric generator that produces a fully labelled staircase point cloud with
realistic sensor artefacts, and details four classical geometric methods that we
implement and compare under an identical 3-class formulation.

\subsection{Problem Formulation and Class Scheme}
We treat staircase perception as a point-wise classification problem. Given an
unorganised cloud $\mathcal{P}=\{\mathbf{x}_i\}_{i=1}^{M}$ with
$\mathbf{x}_i\in\R^3$, the goal is to assign every point a label
$\ell_i\in\{0,1,2\}$, where label $0$ denotes a \emph{tread} (a steppable,
nominally horizontal surface), label $1$ denotes a \emph{riser} (the vertical
face connecting consecutive treads), and label $2$ denotes \emph{other}
(structural edges, sensor noise, and outliers). The tread class is the primary
class of interest, because a robot or assistive device plans footholds on tread
surfaces; consequently we evaluate it in a one-vs-rest setting while still
reporting overall 3-class accuracy.

The decisive geometric cue separating the first two classes is surface
orientation relative to the vertical axis. Let $\nvec$ be the unit surface
normal at a point and $\zhat$ the world vertical. A tread satisfies
$|\nvec^\top\zhat|\to 1$ (its normal is nearly parallel to $\zhat$), whereas a
riser satisfies $|\nvec^\top\zhat|\to 0$ (its normal is nearly orthogonal to
$\zhat$). Every method below exploits this orientation contrast, either
explicitly through a normal-based gate or implicitly through the height
structure of the cloud. Treating orientation as the organising principle yields
a unified pipeline, summarised in Algorithm~\ref{alg:pipeline}, into which each
of the four methods plugs at the clustering or plane-fitting stage.

\begin{algorithm}[t]
\caption{Orientation-aware segmentation (generic)}\label{alg:pipeline}
\begin{algorithmic}[1]
\STATE estimate per-point normals $\nvec_p$ via PCA over $k$ neighbours \eqref{eq:cov}
\STATE split points into horizontal / vertical orientation groups
\STATE cluster or fit planes within each group
\STATE confirm each region by its plane tilt $|\nvec^\top\zhat|$
\STATE label tread / riser / other
\end{algorithmic}
\end{algorithm}

\subsection{Parametric Data Generation}
To obtain a controlled and fully labelled benchmark, we synthesise the staircase
analytically from physical parameters: width $w=2.5$\,m, tread depth
$d=0.30$\,m, step height $h=0.18$\,m, and $N=4$ steps. For step index
$i\in\{0,\dots,N-1\}$, tread points are sampled uniformly over the horizontal
extent of step $i$ and placed at height $ih$ with additive Gaussian height
noise, as in \eqref{eq:tread}:

\begin{equation}\label{eq:tread}
\begin{cases}
x \sim \mathcal{U}(0,\,w) \\
y \sim \mathcal{U}(i d,\,(i{+}1) d) \\
z = i h + \epsilon_z,\quad \epsilon_z \sim \mathcal{N}(0,\sigma_z^2)
\end{cases}
\end{equation}
\begin{equation}\label{eq:riser}
\begin{cases}
x \sim \mathcal{U}(0,\,w) \\
y = i d + \epsilon_y,\quad \epsilon_y \sim \mathcal{N}(0,\sigma_y^2) \\
z \sim \mathcal{U}((i{-}1) h,\, i h)
\end{cases}
\end{equation}

Riser points, generated by \eqref{eq:riser} for $i\in\{1,\dots,N-1\}$ (the
vertical face rising to tread $i$), share the same lateral spread but
lie on a vertical plane at depth $id$ with small along-depth jitter
$\sigma_y=0.02$\,m, their height sweeping uniformly across one step rise. The
nominal height noise is $\sigma_z=0.02$\,m, a value we later vary to probe
sensitivity. On top of this structured cloud we inject realism in two ways.
First, approximately $4\%$ of the points are replaced by uniformly distributed
outliers spanning the staircase bounding volume, mimicking spurious returns and
edge artefacts; these are labelled as \emph{other}. Second, every retained point
receives an isotropic LiDAR-like jitter drawn from $\mathcal{N}(0,0.005^2)$ on
each coordinate, approximating the few-millimetre range noise typical of
indoor depth sensors. We treat this value as a nominal modelling choice rather
than a calibrated sensor model, and probe its principal axis ($\sigma_z$)
directly in Section~\ref{sec:evaluation}. The resulting
cloud contains $18{,}333$ points, partitioned as $12{,}800$ tread, $4{,}800$
riser, and $733$ other. Because the generator emits per-point labels by
construction, it provides exact ground truth for both training-free evaluation
and the noise sweep, removing the manual-annotation bottleneck that hampers
learned segmenters.

\subsection{RANSAC Plane Fitting}
The first method fits planes by Random Sample Consensus. A candidate plane with
unit normal $\nvec$ and offset $b$ scores points by their orthogonal distance,
and the inlier set $\mathcal{I}$ collects points closer than a tolerance $\tau$:

\begin{equation}\label{eq:plane}
d(\mathbf{x}) = |\,\nvec^\top \mathbf{x} + b\,|,\qquad
\mathcal{I} = \{\mathbf{x} : d(\mathbf{x}) < \tau\}
\end{equation}

A naive single fit fails on stairs: the dominant plane that maximises inliers is
the oblique surface spanning the entire staircase, whose normal has
$|\nvec^\top\zhat| = d/\sqrt{d^2+h^2}\approx 0.86$. RANSAC therefore returns a spurious ``ramp''
that absorbs both treads and risers into one steppable region. To prevent this,
we apply a two-phase, orientation-gated scheme: we first extract horizontal
planes (treads) by accepting only fits whose normal satisfies
$|\nvec^\top\zhat|\ge 0.93$, then re-run RANSAC on the residual points to extract
vertical planes (risers) with $|\nvec^\top\zhat|\le 0.25$. We set $\tau=0.035$\,m
and impose minimum inlier counts of $2{,}000$ points for tread planes and
$300$ for riser planes, which suppresses thin riser slices from
contaminating tread fits at the cost of occasionally rejecting sparse vertical
faces. This orientation-aware variant follows the spirit of structure-guided
RANSAC refinements~\cite{li2017improved,jalal2021rgb}.

\subsection{PCA Normal Analysis}
The second method classifies each point purely from its local surface
orientation. For point $p$ we gather its $k$ nearest neighbours
$\mathcal{N}_k(p)$, form the local covariance matrix $\mathbf{C}_p$, and take the
eigenvector associated with its smallest eigenvalue as the surface normal:

\begin{equation}\label{eq:cov}
\mathbf{C}_p = \frac{1}{k}\sum_{j\in\mathcal{N}_k(p)}
(\mathbf{x}_j-\bar{\mathbf{x}})(\mathbf{x}_j-\bar{\mathbf{x}})^\top,\quad
\nvec_p = \mathbf{v}_{\min}(\mathbf{C}_p)
\end{equation}

The smallest-eigenvalue direction is the axis of least point spread, which for a
locally planar patch coincides with its normal~\cite{huang2009consolidation,castillo2013point}.
With $k=50$ the estimate is stable against the injected jitter while remaining
local enough to respect tread--riser boundaries. We classify by the verticality
of the estimated normal: a point is labelled tread when
$|\nvec_p^\top\zhat|\ge 0.80$ and riser when $|\nvec_p^\top\zhat|\le 0.50$;
intermediate values, which arise mostly at edges and from outliers, fall to the
\emph{other} class. This decision is per-point and parameter-light, requiring no
plane-fitting iterations. These PCA thresholds ($0.80/0.50$) are deliberately
looser than the RANSAC orientation gate ($0.93/0.25$): RANSAC tests a global
plane whose aggregate normal is precise, whereas per-point PCA normals are
noisier and require a wider acceptance band.

\subsection{DBSCAN with Slope Confirmation}
The third method couples orientation reasoning with spatial clustering. We first
pre-split the cloud into a horizontal group and a vertical group using the same
normal-verticality criterion as above. Within each group we then run
density-based spatial clustering (DBSCAN) with neighbourhood radius
$\varepsilon=0.08$\,m, so that each physically separate surface forms its own
connected cluster while sparse outliers, lacking dense support, are assigned to
the noise label. Pre-splitting by orientation is essential: it prevents a tread
and the riser directly beneath it---which are spatially adjacent---from merging
into a single density-connected cluster. Each resulting cluster is then
confirmed by its aggregate slope, i.e.\ the verticality $|\nvec^\top\zhat|$ of
its fitted plane, and labelled tread, riser, or other accordingly. This
combination of clustering and region-level geometric confirmation mirrors
established point-cloud clustering and region-growing
practice~\cite{zhou2024identification,luo2021supervoxel}.

\subsection{Height-Histogram Analysis}
The fourth method is the most specialised to the staircase structure and ignores
normals entirely. Because every tread of an ideal staircase lies at a discrete
height $ih$, the marginal distribution of the $z$ coordinate concentrates into
sharp peaks at those levels, separated by low-density bands corresponding to the
riser sweeps. We therefore build a one-dimensional histogram of $z$ over the
whole cloud and detect its dominant peaks; the points falling within a
window of half-width $0.05$\,m around each peak are labelled tread, while points in the intervening
height bands are labelled riser, and the remainder become other. On the
benchmark cloud this recovers four tread levels at $z\approx\{0.00, 0.18, 0.36,
0.54\}$\,m, matching the generative parameters. The method is extremely fast and
requires no neighbourhood computation, but it presumes a regular, axis-aligned
staircase and, as the evaluation shows, its height bands tend to over-include
riser points near each tread level.

\subsection{Implementation and Parameters}
All methods were implemented in Python with NumPy and scikit-learn; RANSAC and
the normal-based gates are custom, while DBSCAN and the nearest-neighbour search
reuse standard library routines. Table~\ref{tab:params} consolidates every
parameter for reproducibility. The four methods share an asymptotic cost
dominated by the $k$-nearest-neighbour normal estimation
($O(M\log M)$ with a $k$-d tree for $M$ points); RANSAC adds a term linear in
the iteration budget, DBSCAN a near-linear neighbourhood query, and the
height-histogram is $O(M)$, making it by far the cheapest. All values were held
fixed across the experiments reported in Section~\ref{sec:evaluation}.

\begin{table}[t]
\caption{Hyper-parameters held fixed across all experiments.}
\label{tab:params}
\centering
\footnotesize
\begin{tabular}{lll}
\toprule
Symbol & Value & Description \\
\midrule
$w,d,h$ & $2.5,\,0.30,\,0.18$\,m & width, tread depth, step height \\
$N$ & $4$ & number of steps \\
$\sigma_z,\sigma_y$ & $0.02,\,0.02$\,m & tread / riser Gaussian noise \\
jitter & $0.005$\,m & isotropic LiDAR-like jitter \\
outliers & $4\%$ & uniform outlier fraction \\
$\tau$ & $0.035$\,m & RANSAC inlier threshold \\
min.\ inliers & $2{,}000\,/\,300$ & RANSAC tread / riser plane \\
$k$ & $50$ & PCA normal neighbours \\
gate (RANSAC) & $0.93\,/\,0.25$ & horiz.\ / vert.\ $|\nvec^\top\zhat|$ \\
gate (PCA) & $0.80\,/\,0.50$ & tread / riser $|\nvec^\top\zhat|$ \\
$\varepsilon$ & $0.08$\,m & DBSCAN radius \\
window & $0.05$\,m & histogram peak half-width \\
\bottomrule
\end{tabular}
\end{table}
